---
title: "Exercise 10"
format: html
editor_options: 
  chunk_output_type: console
---

## Load in required packages
```{r}
#| output: false

library(tidyverse)
library(skimr)
library(broom)
library(mosaic)
library(effectsize)
library(patchwork)
library(emmeans)
library(ggpubr)
```

# Preliminaries
Load in the AVONET dataset.
```{r}
#| output: false
d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv")
```

Winnow the dataset to include only wanted variables.
```{r}
d <- d %>%
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Primary.Lifestyle, Min.Latitude, Max.Latitude, Centroid.Latitude, Range.Size)
```

Do a bit of exploratory data analysis using {skimr}. Which of the variables are categorical and which are numeric?
```{r}
skim(d)
```
**Categorical:** Species, Family, Order, Habitat, Trophic Level, Trophic Niche, and Primary Lifestyle  
**Numeric:** Beak Length, Beak Width, Beak Depth, Tarsus Length, Wing Length, Tail Length, Mass, Migration, Minimum Latitude, Maximum Latitude, Centroid Latitude, and Range Size

# Challenge 1
# One-Factor ANOVA and Inference

## Step 1
Make boxplots of log(Mass) in relation to Trophic.Level and Migration behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, you will want to convert the variable Migration (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable.
```{r}
# convert Migration to a factor
d <- d %>%
  mutate(Migration = as.factor(Migration))

# boxplot of log(Mass) in relation to Trophic.Level
p1 <- ggplot(data = d %>% drop_na(Trophic.Level),
             aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot()+
  geom_jitter(alpha = 0.1)

# boxplot of log(Mass) in relation to Migration
p2 <- ggplot(data = d %>% drop_na(Migration),
             aes(x = Migration, y = log(Mass))) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1)

p1 + p2
```

## Step 2
Run linear models using the lm() function to look at the relationship between log(Mass) and Trophic.Level and between log(Mass) and Migration.
```{r}
m1 <- lm(log(Mass) ~ Trophic.Level, data = d)
m2 <- lm(log(Mass) ~ Migration, data = d)
```

Examine the output of the resultant linear models. Is log(Mass) associated with either Trophic.Level or Migration category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?
```{r}
summary(m1)
summary(m2)
```
The p-value of the F-statistic for both Trophic Level and Migration is significant (i.e., p < 2.2e-16). Therefore, log(Mass) *is* associated with both Trophic Level and Migration category.

Given the regression coefficients returned for your Migration model, which Migration categor(ies) are different than the reference level? What level is the reference level? 

Migration categories 2 ("Partially migratory") and 3 ("Fully migratory") are significantly different than the reference level (category 1, "Sedentary").

Relevel and assess differences among the remaining pair of Migration categories.
```{r}
# relevel
d <- d %>%
  mutate(Migration = relevel(Migration, ref = "3"))

# re-run model
m2 <- lm(log(Mass) ~ Migration, data = d)
summary(m2)
```
After releveling, migration category still has a significant effect on log(Mass) and  migration categories 1 and 2 are significantly different from migration category 3. Therefore, all of the migration categories are significantly different from one another in terms of log(Mass).

## Step 3
Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another.
```{r}
m1aov <- aov(log(Mass) ~ Migration, data = d)
posthoc <- TukeyHSD(m1aov, which = "Migration", conf.level = 0.95)
posthoc
```
The post-hoc test confirms that the Migration categories are all significantly different from one another. 

## Step 4
Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(Mass) in relation to Trophic.Level and calculate a p value for your original F statistic. You can do this either by programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) or by using the {infer} workflow and setting calculate(stat="F").
```{r}
# F statistic and p value for omnibus F test 
original.F <- aov(log(Mass) ~ Trophic.Level, data = d) %>%
  tidy() %>%
  filter(term == "Trophic.Level")
original.F

# ANOVA by permutation
nperm <- 1000
permuted.F <- vector(length = nperm)

for (i in 1:nperm) {
  d_perm <- d
  d_perm$Trophic.Level <- sample(d$Trophic.Level) # shuffle
  permuted.F[[i]] <- aov(log(Mass) ~ Trophic.Level, data = d_perm) %>%
    tidy() %>%
    filter(term == "Trophic.Level") %>%
    pull(statistic)
}
histogram(permuted.F)

# get p-value of original F statistic
p.value <- mean(permuted.F > original.F$statistic)
p.value
```

# Challenge 2
# Data Wrangling plus One- and Two-Factor ANOVA

## Step 1
Create two new variables:

* Relative beak length, which you should calculate as the residual of log(Beak.Length_Culmen) on log(Mass).
* Relative tarsus length, which you should calculate as the residual of log(Tarsus.Length) on log(Mass).
```{r}
# relative beak length
beak.m <- lm(log(Beak.Length_Culmen) ~ log(Mass), data = d)
d$Rel_Beak <- beak.m$residuals

# relative tarsus length
tarsus.m <- lm(log(Tarsus.Length) ~ log(Mass), data = d)
d$Rel_Tarsus <- tarsus.m$residuals
```

## Step 2
Make a boxplot or violin plot of your new relative tarsus length variable in relation to Primary.Lifestyle and of your new relative beak length variable in relation to Trophic.Niche.
```{r}
# relative tarsus length vs. primary lifestyle
p1 <- d %>%
  ggplot(aes(x = factor(Primary.Lifestyle, levels = c("Aerial", "Aquatic", "Insessorial", "Terrestrial", "Generalist")), y = Rel_Tarsus)) +
  geom_boxplot() +
  theme_classic() +
  ylab("Relative Tarsus Length") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

# relative beak length vs. trophic niche
p2 <- d %>%
  drop_na(Trophic.Niche) %>%
  ggplot(aes(x = Trophic.Niche, y = Rel_Beak)) +
  geom_boxplot() +
  theme_classic() +
  ylab("Relative Beak Length") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

p1 + p2
```

## Step 3
Run ANOVA analyses to look at the association between geographic range size and the variable Migration. You should first drop those observations for which Migration is not scored and also look at the distribution of the variable Range.Size to decide whether and how it might need to be transformed. Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with Migration behavior style?
```{r}
# look at distribution of Range Size
histogram(d$Range.Size)
```
Since it is right-skewed, needs to be log transformed

```{r}
# run ANOVA
aov1 <- aov(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration))
summary(aov1)

# effect size
effectsize(aov1)
```
Range size *is* associated with form of migration. 9% of the variance in range size is accounted for by variance in migration category. 

Given the regression coefficients returned in the output of the model, which Migration categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of Migration categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another.
```{r}
# linear model
m3 <- lm(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration))
summary(m3)
```
Migration categories 1 and 2 differ significantly from the reference category (3).

```{r}
# relevel
d <- d %>%
  mutate(Migration = relevel(Migration, ref = "1"))

# re-run model
m3 <- lm(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration))
summary(m3)
```
Migration category 1 also significantly differs from category 3.

```{r}
# post-hoc Tukey test
posthoc <- TukeyHSD(aov1, which = "Migration", conf.level = 0.95)
posthoc
```
All Migration categories differ significantly from one another in terms of range size.

## Step 4
Winnow your original data to just consider birds from the Infraorder “Passeriformes” (song birds).
```{r}
pass <- d %>%
  filter(Order1 == "Passeriformes")
```

Run separate one-factor ANOVA analyses to look at the association between [1] relative beak length and Primary.Lifestyle and between [2] relative beak length and Trophic.Level. In doing so…

* Make boxplots of response variable by each predictor and by the combination of predictors.
* Run linear models for each predictor separately and interpret the model output.
```{r}
#| fig-height: 8
# one-way ANOVAs
aov1 <- aov(Rel_Beak ~ Primary.Lifestyle, data = pass)
summary(aov1)
aov2 <- aov(Rel_Beak ~ Trophic.Level, data = pass)
summary(aov2)

# boxplots of response vs. predictor and by the combination of predictors
p1 <- ggplot(data = d,
             aes(x = Primary.Lifestyle, y = Rel_Beak)) +
  geom_boxplot()+
  geom_jitter(alpha = 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

p2 <- ggplot(data = d %>% drop_na(Trophic.Level),
             aes(x = Trophic.Level, y = Rel_Beak)) +
  geom_boxplot()+
  geom_jitter(alpha = 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

p3 <- ggplot(data = d %>% drop_na(Trophic.Level),
             aes(x = Primary.Lifestyle, y = Rel_Beak)) +
  geom_boxplot()+
  geom_jitter(alpha = 0.1) +
  facet_wrap(~Trophic.Level) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

p4 <- ggplot(data = d %>% drop_na(Trophic.Level),
             aes(x = Trophic.Level, y = Rel_Beak)) +
  geom_boxplot()+
  geom_jitter(alpha = 0.1) +
  facet_wrap(~Primary.Lifestyle) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

(p1 + p2) / p3 / p4 + plot_layout(heights = c(1,2,2))

# run linear models for each predictor separately
lm1 <- lm(Rel_Beak ~ Primary.Lifestyle, data = pass)
lm2 <- lm(Rel_Beak ~ Trophic.Level, data = pass)
summary(lm1)
summary(lm2)
```
Both primary lifestyle and trophic level have a significant effect on relative beak length.

## Step 5
Run a two-factor model to look at the association between relative beak length and both Primary.Lifestyle and Trophic.Level among the passeriforms. Based on the model output, what would you conclude about how relative beak length is related to these two variables?
```{r}
aov3 <- aov(Rel_Beak ~ Primary.Lifestyle + Trophic.Level, data = pass)
summary(aov3)
```
Both primary lifestyle and trophic level are significant predictors of relative beak length.

## Step 6
Run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term. To do this, you should modify your model formula using the colon operator (:) to specify the interaction, e.g., relative beak length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level. Based on the model output, what would you now conclude about how relative beak length is related to these two variables?
```{r}
aov4 <- aov(Rel_Beak ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = pass)
summary(aov4)
```
There is an interaction effect of primary lifestyle and trophic level. Therefore, primary lifestyle, trophic level, and the interaction between primary lifestyle and trophic level are significant predictors of relative beak length. 

## Step 7
Use the interaction.plot() and either the plot_model() or emmip() function to visualize the interaction between Primary.Lifestyle and Trophic.Level, based on both [1] the raw dataset and [2] your model that includes an interaction term.
```{r}
#| warning: false

# raw dataset
interaction.plot(x.factor = factor(pass$Primary.Lifestyle),
                 xlab = "Primary Lifestyle",
                 trace.factor = pass$Trophic.Level,
                 trace.label = "Trophic Level",
                 response = pass$Rel_Beak,
                 ylab = "Relative Beak Length",
                 fun = base::mean)

interaction.plot(x.factor = factor(pass$Trophic.Level),
                 xlab = "Trophic Level",
                 trace.factor = pass$Primary.Lifestyle,
                 trace.label = "Primary Lifestyle",
                 response = pass$Rel_Beak,
                 ylab = "Relative Beak Length",
                 fun = base::mean)

# model with interaction term
emmip(aov4,
      Trophic.Level ~ Primary.Lifestyle,
      CIs = TRUE)

emmip(aov4,
      Primary.Lifestyle ~ Trophic.Level,
      CIs = TRUE)
```

## Step 8
See whether variances in across groups in your various models (e.g., for relative beak length ~ trophic level) are roughly equal. Additionally, do a visual check of whether observations and model residuals within groups look to be normally distributed.
```{r}
#| message: false
#| warning: false

## m1 <- lm(log(Mass) ~ Trophic.Level, data = d)

# check equal variance 
stats <- d %>%
  group_by(Trophic.Level) %>%
  summarise("sd(log(Mass))" = sd(log(Mass)))
max(stats$`sd(log(Mass))`) / min(stats$`sd(log(Mass))`)

# check normal distribution of observations
p1 <- ggplot(data = d %>% drop_na(Trophic.Level), 
             aes(x = log(Mass))) +
  geom_histogram() +
  facet_wrap(~Trophic.Level)

# check normal distribution of residuals within groups
m1_res <- tibble("residuals" = m1$residuals, "trophic.level" = m1$model$Trophic.Level)
p2 <- ggqqplot(data = m1_res,
               x = "residuals") +
  facet_wrap(~trophic.level)

p1+p2


## m2 <- lm(log(Mass) ~ Migration, data = d)

# check equal variance 
stats <- d %>%
  drop_na(Migration) %>%
  group_by(Migration) %>%
  summarise("sd(log(Mass))" = sd(log(Mass)))  
max(stats$`sd(log(Mass))`) / min(stats$`sd(log(Mass))`)

# check normal distribution of observations
p3 <- ggplot(data = d %>% drop_na(Migration),
             aes(x = log(Mass))) +
  geom_histogram() +
  facet_wrap(~Migration)

# check normal distribution of residuals within groups
m2_res <- tibble("residuals" = m2$residuals, "migration" = m2$model$Migration)
p4 <- ggqqplot(data = m2_res,
               x = "residuals") +
  facet_wrap(~migration)

p3+p4


## m3 <- lm(log(Range.Size) ~ Migration, data = d %>% drop_na(Migration))

# check equal variance 
stats <- d %>%
  drop_na(Migration) %>%
  drop_na(Range.Size) %>%
  group_by(Migration) %>%
  summarise("sd(log(Range.Size))" = sd(log(Range.Size)))  
max(stats$`sd(log(Range.Size))`) / min(stats$`sd(log(Range.Size))`)

# check normal distribution of observations
p5 <- ggplot(data = d %>% drop_na(Migration),
             aes(x = log(Range.Size))) +
  geom_histogram() +
  facet_wrap(~Migration)

# check normal distribution of residuals within groups
m3_res <- tibble("residuals" = m3$residuals, "migration" = m3$model$Migration)
p6 <- ggqqplot(data = m3_res,
               x = "residuals") +
  facet_wrap(~migration)

p5+p6


## lm1 <- lm(Rel_Beak ~ Primary.Lifestyle, data = pass)

# check equal variance 
stats <- pass %>%
  group_by(Trophic.Level) %>%
  summarise("mean(Rel_Beak)" = mean(Rel_Beak),
            "sd(Rel_Beak" = sd(Rel_Beak))  
max(stats$`sd(Rel_Beak`) / min(stats$`sd(Rel_Beak`)

# check normal distribution of observations
p7 <- ggplot(data = pass,
             aes(x = Rel_Beak)) +
  geom_histogram() +
  facet_wrap(~Trophic.Level)

# check normal distribution of residuals within groups
lm2_res <- tibble("residuals" = lm2$residuals, "trophic.level" = lm2$model$Trophic.Level)
p8 <- ggqqplot(data = lm2_res,
               x = "residuals") +
  facet_wrap(~trophic.level)

p7+p8


## lm2 <- lm(Rel_Beak ~ Trophic.Level, data = pass)

# check equal variance 
stats <- pass %>%
  group_by(Primary.Lifestyle) %>%
  summarise("mean(Rel_Beak)" = mean(Rel_Beak),
            "sd(Rel_Beak" = sd(Rel_Beak))  
max(stats$`sd(Rel_Beak`) / min(stats$`sd(Rel_Beak`)

# check normal distribution of observations
p9 <- ggplot(data = pass,
             aes(x = Rel_Beak)) +
  geom_histogram() +
  facet_wrap(~Primary.Lifestyle)

# check normal distribution of residuals within groups
lm1_res <- tibble("residuals" = lm1$residuals, "primary.lifestyle" = lm1$model$Primary.Lifestyle)
p10 <- ggqqplot(data = lm1_res,
               x = "residuals") +
  facet_wrap(~primary.lifestyle)

p9+p10
```
The ratios of max/min is < 2 for most of the models (except log(Mass) ~ Trophic.Level), so equal variance is reasonable for most of them. However, normality of both the observations and the residuals is not really met for all of the models, especially for the residuals. Therefore, we are not really supposed to be using linear regression/ANOVAs to analyze these data.
